---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Isometric height maps

## Intro
* This readme documents a project I started nearly 1.5 years ago and only revisited again in the last week!
* It's an attempt at using isometric projection to create pseudo 3D images from data. I'm sharing it here primarily to document what I learnt but also in case it can be a useful starting point for someone else.
* This is a project that I created for fun and personal development, so its slow and likely to be inaccurate. **It's certainly experimental**. There are however some incredibly cool R packages for doing this type of thing properly...
  * [rayshader](https://github.com/tylermorganwall/rayshader)
  * [isocubes](https://github.com/coolbutuseless/isocubes)

##### Disclaimer
* I am not a smart man.
  * My maths is...OK, but I have very much 'brute force' got this to work via naive copying of equations from [this  Wikipedia article](https://en.wikipedia.org/wiki/Isometric_projection). My test of *'is it working?'* is based solely on a visual inspection of the outputs to see that they don't look obviously wrong!
  * As a result, I may have got things incorrect, or had to insert spurious minus signs to get things to not render as mirror images of themselves etc!
  
## Packages
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(magick)
```

## Rotation matrices and unit-cube
* Define a function `iso_project` that takes
   1. The rotation angles around the vertical and horizontal axes (in degrees because that's how my brain works)
   2. The 3D coordinates x, y, z
* This function then computes the rotation matrices for these angles and applies the matrix multiplication transformation (vectorised) over x, y and z 
* This is my interpretation of the [Wikipedia article](https://en.wikipedia.org/wiki/Isometric_projection)
```{r}
iso_project <-
  function(x, y, z,
           va = 45, 
           ha = asin(tan(pi/6))*(180/pi)
  ){
    
    # Convert rotation angles to radians
    var <- va*(pi/180)
    har <- ha*(pi/180)
    
    # Construct rotation matrices
    m1a <- 
      matrix(c(1, 0, 0,
               0, cos(har), sin(har),
               0, -sin(har), cos(har)),
             byrow = TRUE,
             ncol = 3)
    
    m1b <- 
      matrix(c(cos(var), 0, -sin(var),
               0, 1, 0,
               sin(var), 0, cos(var)),
             byrow = TRUE,
             ncol = 3)
    
    m2 <- matrix(c(1,0,0,0,1,0,0,0,0), byrow = TRUE, nrow=3, ncol=3)
    
    # Perform vectorised matrix multiplication
    Vectorize(function(x, y, z) m2 %*% ((m1a %*% m1b) %*% matrix(c(x, y, z))))(x, y, z)
  }
```

* Now define the x, y, z coordinates of the 3 sides of the unit cube that we will be able to see
```{r}
unit_cube_df <-
  tribble(
    ~face,   ~ux, ~uy, ~uz,
    "right", 0,   0,   0,
    "right", 0,   1,   0,
    "right", 1,   1,   0,
    "right", 1,   0,   0,
    "left",  0,   0,   0,
    "left",  0,   0,   1,
    "left",  0,   1,   1,
    "left",  0,   1,   0,
    "top",   0,   1,   0,
    "top",   1,   1,   0,
    "top",   1,   1,   1,
    "top",   0,   1,   1)
```

* Use `iso_project` to transform the coordinates of the unit cube and then visualise each cube face in the projected space
```{r, fig.width=8, fig.height=2.5}
# Transform the unit cube x y z coordinates
uc <- 
  unit_cube_df %>% 
  mutate(iso_project(ux, uy, uz) %>% 
           t() %>% 
           as_tibble(.name_repair = ~c("x", "y", "z")))

# Visualise the cube faces
ggplot()+
  geom_polygon(data = uc, aes(x, y, group = face, fill = face))+
  geom_polygon(data = uc %>% rename(face2=face), aes(x, y, group = face2), col = 1, fill=NA)+
  coord_equal()+
  facet_wrap(~face, nrow=1)+
  theme(legend.position = "")
```


## Volcano
Now let's try this on some real topographic data, using the R `volcano` dataset.

> Maunga Whau (Mt Eden) is one of about 50 volcanos in the Auckland volcanic field. This data set gives topographic information for Maunga Whau on a 10m by 10m grid

* Scale the data to start from 0 and divide by 10 to roughly get the height values to be proportional to the number of columns and rows in the matrix
```{r}
v <- as.vector(volcano - min(volcano))/10
```

* Construct a viridis colour scale that maps to the height values
* Define how much shading to apply to the left and right side of the cubes
```{r}
cols <- 
  colorRamp(viridis::turbo(n_distinct(v)))(scales::rescale(v, c(0,1))) %>% 
  apply(1, function(x) rgb(x[1], x[2], x[3], maxColorValue = 255))

# Define factor for how much to lighten/darken the sides of the cuboids
f <- 0.25
```

* All of the processing is done in this next step
  * Construct a dataframe of the x-z coordinates for the volcano data set
  * Add the height values y, the colours and create a unique block_id per row
  * Merge the unit cube dataframe into each row (making each row duplicate 12 times)
  * Shift the unit cube in x and z and stretch/squash it's height in y
  * Perform the isometric projection
  * Convert the colours to HSV and then darken the right and left sides by reducing V 
  * Order the dataframe to plot in the correct order (so polygons overlap each other correctly)
  * Plot!
```{r, dpi=600, fig.width=15, fig.height=9}
expand.grid(
  z = (nrow(volcano)-1):0,
  x = 0:(ncol(volcano)-1)) %>% 
  mutate(y = v, 
         col = cols, 
         block_id = row_number()) %>% 
  merge(unit_cube_df, all = TRUE) %>%
  mutate(ux = ux+x, 
         uy = uy*y, 
         uz = uz+z,
         iso_project(ux, uy, uz) %>% t() %>% as_tibble(.name_repair = ~c("px", "py", "pz")),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         v2 = case_when(face == "top" ~ v,
                        face == "right" ~ v*(1-f),
                        face == "left" ~ v*(1-(2*f))),
         col2 = hsv(h, s, v2),
         my_group = paste0(face, block_id)) %>%
  arrange(desc(x), desc(z)) %>% 
  mutate(my_group = fct_inorder(my_group)) %>% 
  ggplot(aes(px, py, fill=I(col2), group=my_group))+
  geom_polygon(col=NA)+
  coord_equal()

ggsave("volcano.pdf", width=30, height=17)
```

* Run on `volcano` again
  * Re-define the colour scheme
  * Highlight the highest point in red
  * Highlight a range of points in orange 

```{r, dpi=600, fig.width=15, fig.height=9}
cols <- 
  colorRamp(viridis::mako(n_distinct(v)))(scales::rescale(v, c(0,1))) %>% 
  apply(1, function(x) rgb(x[1], x[2], x[3], maxColorValue = 255))

# Highlight highest part in red
cols[which(v == max(v))] <- "red"

# Highlight areas between 4 and 5
cols[between(v, 5, 6)] <- "orange"

expand.grid(
  z = (nrow(volcano)-1):0,
  x = 0:(ncol(volcano)-1)) %>% 
  mutate(y = v, 
         col = cols, 
         block_id = row_number()) %>% 
  merge(unit_cube_df, all = TRUE) %>%
  mutate(ux = ux+x, 
         uy = uy*y, 
         uz = uz+z,
         iso_project(ux, uy, uz) %>% t() %>% as_tibble(.name_repair = ~c("px", "py", "pz")),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         v2 = case_when(face == "top" ~ v,
                        face == "right" ~ v*(1-f),
                        face == "left" ~ v*(1-(2*f))),
         col2 = hsv(h, s, v2),
         my_group = paste0(face, block_id)) %>%
  arrange(desc(x), desc(z)) %>% 
  mutate(my_group = fct_inorder(my_group)) %>% 
  ggplot(aes(px, py, fill=I(col2), group=my_group))+
  geom_polygon(col=NA)+
  coord_equal()

ggsave("volcano-highlights.pdf", width=30, height=17)
```

## Images
### Sweets
* Now try the code on an image rather than topographic data
  * Read and perform some processing on an image
  * Convert the image to a dataframe
  * Add a pseudo height column that is the brightness (v) from the hsv colour of each pixel
  * Here I have arbitrarily scaled the height to between 0 and 20
  * Then perform the same projection as above and visualise both outputs
```{r, dpi=600, fig.width=15, fig.height=6}
# Define horizontal image size in pixels
s <- 200

# Define factor for how much to lighten/darken the sides of the cuboids
f <- 0.2

img_df <- 
  image_read('https://cdn.wallpapersafari.com/89/79/wW6LRs.jpg') %>% 
  image_scale(geometry = paste0(s,"x",s,"^")) %>%
  image_crop(geometry = paste0(s,"x",s), gravity = "center") %>%
  image_flip() %>% 
  image_blur() %>%
  image_raster() %>%
  rename(z = y) %>% 
  mutate(x = x-1,
         z = z-1,
         col2rgb(col) %>% t() %>% as_tibble(),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         y = scales::rescale(v, c(0,20)),
         block_id = row_number()) %>% 
  merge(unit_cube_df, all = TRUE) %>% 
  mutate(ux = ux+x, uy = uy*y, uz = uz+z)
         
out <-
  img_df %>% 
  mutate(iso_project(ux, uy, uz) %>% t() %>% as_tibble(.name_repair = ~c("px", "py", "pz")),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         v2 = case_when(face == "top" ~ v,
                        face == "right" ~ v*(1-f),
                        face == "left" ~ v*(1-(2*f))),
         col2 = hsv(h, s, v2),
         my_group = paste0(face, block_id)) %>%
  arrange(desc(x), desc(z)) %>% 
  mutate(my_group = fct_inorder(my_group)) %>% 
  ggplot(aes(px, py, fill=I(col2), group=my_group))+
  geom_polygon(col=NA)+
  coord_equal()

patchwork::wrap_plots(
  img_df %>% ggplot() + geom_raster(aes(x, z, fill = I(col))) + coord_equal(),
  out
)

ggsave("sweets.pdf", width=30, height=15)
```

* Of course, this is completely fake 3D! I have no information about the height of the sweets - I have made it up
* As brighter pixels are shown to be 'higher', it sort of looks not bad, but just a bit off
  * Note that the darker purple sweets are much less pronounced, but the white reflections off their surface are very high!

### Water
* Another example
```{r, dpi=600, fig.width=15, fig.height=6}
# Define horizontal image size in pixels
s <- 200

# Define factor for how much to lighten/darken the sides of the cuboids
f <- 0.2

img_df <- 
  image_read('https://kaboompics.com/cache/8/6/f/5/f/86f5fef3f896cf25c35fac3a3effd557e87fd4c9.jpeg') %>% 
  image_scale(geometry = paste0(s,"x",s,"^")) %>%
  image_crop(geometry = paste0(s,"x",s), gravity = "center") %>%
  image_flip() %>% 
  image_blur(sigma = 1) %>%
  image_raster() %>%
  rename(z = y) %>% 
  mutate(x = x-1,
         z = z-1,
         col2rgb(col) %>% t() %>% as_tibble(),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         y = scales::rescale(v, c(0, 5)),
         block_id = row_number()) %>% 
  merge(unit_cube_df, all = TRUE) %>% 
  mutate(ux = ux+x, uy = uy*y, uz = uz+z)
         
out <-
  img_df %>% 
  mutate(iso_project(ux, uy, uz) %>% t() %>% as_tibble(.name_repair = ~c("px", "py", "pz")),
         col2rgb(col) %>% rgb2hsv() %>% t() %>% as_tibble(),
         v2 = case_when(face == "top" ~ v,
                        face == "right" ~ v*(1-f),
                        face == "left" ~ v*(1-(2*f))),
         col2 = hsv(h, s, v2),
         my_group = paste0(face, block_id)) %>%
  arrange(desc(x), desc(z)) %>% 
  mutate(my_group = fct_inorder(my_group)) %>% 
  ggplot(aes(px, py, fill=I(col2), group=my_group))+
  geom_polygon(col=NA)+
  coord_equal()

patchwork::wrap_plots(
  img_df %>% ggplot() + geom_raster(aes(x, z, fill = I(col))) + coord_equal(),
  out
)

ggsave("water.pdf", width=30, height=15)
```


# Thoughts
* This wont work well with negative height values